---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
Fabian Mair
04/15/20

```{r}
library(ggplot2)
library(dplyr)
if(!file.exists("./data")){dir.create("./data")}
fileUrl <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
download.file(fileUrl,destfile="./data/activity.zip",method="curl")

unzip(zipfile="./data/activity.zip",exdir="./data")
activity <- read.csv("./data/activity.csv")
activity$date <- as.Date(activity$date)

```

# Steps per day, first 10 rows
```{r}
stepsPerDay <- activity %>%
        group_by(date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 
head(stepsPerDay,10)
```

## What is mean total number of steps taken per day?
Calculating mean and median
```{r}
meanPreNA <- round(mean(stepsPerDay$sumsteps),digits = 2)
medianPreNA <- round(median(stepsPerDay$sumsteps),digits = 2)
```

The mean
```{r}
print(paste("The mean is: ", meanPreNA))
```
The median
```{r}
print(paste("The median is: ", medianPreNA))
```

I genereated the plot
```{r}
hist(stepsPerDay$sumsteps, main = "Histogram of Daily Steps", 
     col="gray", xlab="Steps", ylim = c(0,30))
```

# Steps per Interval, first 10 rows
```{r}
stepsPerInterval <- activity %>%
        group_by(interval) %>%
        summarize(meansteps = mean(steps, na.rm = TRUE)) 
#Display first 10 rows of data
head(stepsPerInterval,10)
```

Interval containing the most steps on average
```{r}
print(paste("Interval containing the most steps on average: ",stepsPerInterval$interval[which.max(stepsPerInterval$meansteps)]))
```

Average steps for that interval
```{r}
print(paste("Average steps for that interval: ",round(max(stepsPerInterval$meansteps),digits=2)))
```

I genereated the plot
```{r}
plot(stepsPerInterval$meansteps ~ stepsPerInterval$interval,
     col="black", type="l", xlab = "Intervals (5 Minute)", ylab = "Average Number of Steps",
     main = "Steps By Time Interval")
```


## Imputing missing values
```{r}
print(paste("The total number of rows with NA is: ",sum(is.na(activity$steps))))
head(activity,10)
```


The NA strategy 
```{r}
activityNoNA <- activity  
for (i in 1:nrow(activity)){
        if(is.na(activity$steps[i])){
                activityNoNA$steps[i]<- stepsPerInterval$meansteps[activityNoNA$interval[i] == stepsPerInterval$interval]
        }
}
head(activityNoNA,10)
```

Mean and Median calculation
```{r}
meanPostNA <- round(mean(stepsPerDay$sumsteps), digits = 2)
medianPostNA <- round(median(stepsPerDay$sumsteps), digits = 2)
```

Mean
```{r}
print(paste("The mean is: ", mean(meanPostNA)))
```

Median
```{r}
print(paste("The median is: ", median(medianPostNA)))
```

Comparison
```{r}
NACompare <- data.frame(mean = c(meanPreNA,meanPostNA),median = c(medianPreNA,medianPostNA))
rownames(NACompare) <- c("Pre NA Transformation", "Post NA Transformation")
print(NACompare)
```

I genereated the plot
```{r}
stepsPerDay <- activityNoNA %>%
        group_by(date) %>%
        summarize(sumsteps = sum(steps, na.rm = TRUE)) 
head(stepsPerDay,10)
```

```{r}
hist(stepsPerDay$sumsteps, main = "Histogram of Daily Steps", 
     col="gray", xlab="Steps")
```

## Are there differences in activity patterns between weekdays and weekends?
# Acitivity patterns on weekdays and weekends 
creating new factors indicating whether a given date is a weekday or weekend day.
```{r}
activityDoW <- activityNoNA
activityDoW$date <- as.Date(activityDoW$date)
activityDoW$day <- ifelse(weekdays(activityDoW$date) %in% c("Saturday", "Sunday"), "weekend", "weekday")
activityDoW$day <- as.factor(activityDoW$day)

```

```{r}
activityWeekday <- filter(activityDoW, activityDoW$day == "weekday")
activityWeekend <- filter(activityDoW, activityDoW$day == "weekend")

activityWeekday <- activityWeekday %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
activityWeekday$day <- "weekday"

activityWeekend <- activityWeekend %>%
        group_by(interval) %>%
        summarize(steps = mean(steps)) 
activityWeekend$day <- "weekend"

wkdayWkend <- rbind(activityWeekday, activityWeekend)
wkdayWkend$day <- as.factor(wkdayWkend$day)

```

I genereated the plot
```{r}
g <- ggplot (wkdayWkend, aes (interval, steps))
g + geom_line() + facet_grid (day~.) + 
        theme(axis.text = element_text(size = 14),axis.title = element_text(size = 16)) + 
        labs(y = "Number of Steps") + labs(x = "Interval") + 
        ggtitle("Average Number of Steps - Weekday vs. Weekend") + 
        theme(plot.title = element_text(hjust = 0.5))
```


There is a little difference between the weekends and weekdays. 